Measuring contrast processing in the visual system using SSVEP (and VEP?)

### 2024-03-29 review.qmd
tidied
Some edits on square / sine / harmonics

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for at least 100 years. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high SNR and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years[@Dawson1954]. Because our thinking about SSVEPs is best described in simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This document can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

@@

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal. The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker produces a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Variants involving square wave flicker can additionally produce more complex spectral components. The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. SSVEP signals can also be elicited by periodic changes of stimulus properties other than contrast, such as chromaticity, motion, stereo depth, and facial identity or expression [see @Norcia2015, for an overview]; however our focus here is on the contrast response.

### Why measure responses to contrast?
Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over time ('spatial contrast') or space ('spatial contrast'). Cone photoreceptors contribute to both chromatic and achromatic contrast and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well. [McKeefry, Sutter, Baseler, DiRusso...].

Contrast is relatively simple to define: typically, contrast is specificed as the percentage deviation of uniform stimulus from the background. So, for example, a disk of 100 units, $I_{\mathrm{stim}}$ of cone activation surrounded by a 'background' of 50 units of activation $I_{\mathrm{background}}$ has a contrast of $\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}$ = 100%. Where patterns are more complex (for example, the sine-wave gratings or Gabor patches common in vision science), the Michaelson definition of contrast is specified by the maximum and minimum excursions from the mean: $${{\frac {I_{\mathrm {stimmax} }-I_{\mathrm {stimmin} }}{I_{\mathrm {stimmax} }+I_{\mathrm {stimmin} }}},} $$. These contrast definitions are appropriate both to photometric measures of stimulus contrast (for example, luminance [LenniePokornySmith]) and also to definitions based on cone excitations [DKL, MacBoynton] which are more common in work on chromatic processing.

>>>>

Contrast is the currency of the early visual system. Measuring the way that the computations underlying contrast processing depend on factors such as spatial and temporal frequency, age, eccentricity, chromaticity and the presence of other stimuli has been a focus of vision science for at least 100 years. One of the most productive experimental approaches in this field has been the use of the 'steady-state visually-evoked potential' (SSVEP): a technique where contrast modulating inputs are 'frequency tagged (presented at well-defined frequencies and phases) and the electrical signals that they generate in the brain are analyzed in the temporal frequency domain. SSVEPs have several advantages over conventional measures of visually-evoked responses: they have relatively unambiguous ouput measures, a high SNR and they allow us to analyze interactions between stimulus components using a convenient mathematical framework. Here we describe how SSVEPs have been used to study visual contrast over the past 70 years[@Dawson1954]. Because our thinking about SSVEPs is well-described by simple mathematical models, we embed code that illustrates key steps in the modelling and analysis. This document can therefore be used both as a review of the use of SSVEP in measuring human contrast processing, and as an interactive learning aid.

@@

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal. The flickering stimulus entrains neural responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. SSVEP signals can also be elicited by periodic changes of stimulus properties other than achromatic and chromatic contrast, such as, motion, stereo depth, and facial identity or expression [see @Norcia2015, for an overview]; however our focus here is on the contrast response.

### Why measure responses to contrast?
Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over time ('spatial contrast') or space ('spatial contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well. [McKeefry, Sutter, Baseler, DiRusso...].

Contrast is relatively simple to define: typically, contrast is specified as the percentage deviation of uniform stimulus from the background. So, for example, a disk of 100 units, $I_{\mathrm{stim}}$ of cone activation surrounded by a 'background' of 50 units of activation $I_{\mathrm{background}}$ has a contrast of $\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}$ = 100%. Where patterns are more complex (for example, the sine-wave gratings or Gabor patches common in vision science), the Michaelson definition of contrast is specified by the maximum and minimum excursions from the mean: $${{\frac {I_{\mathrm {stimmax} }-I_{\mathrm {stimmin} }}{I_{\mathrm {stimmax} }+I_{\mathrm {stimmin} }}},} $$. These contrast definitions are appropriate both to photometric measures of stimulus contrast (for example, luminance [LenniePokornySmith]) and also to definitions based on cone excitations [DKL, MacBoynton] which are more common in work on chromatic processing.
